---
title: "Microbiome Data Analysis Using Biclustering Approaches - Bibit"
author: "Italo Rossi del Aguila"
format: pdf
editor: visual
---

```{r libraries}
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)
library(stringr)
```

# Data Set

### Load data

During import, transpose to place samples on the rows and taxa on the columns. Load metadata for labeling and coloring plots.

```{r data_load, include=FALSE}
#| echo: false
#| message: false
#| warning: false
X <- read.delim("data/microbiome_dataSet1.tsv",row.names=1)
X <- t(as.matrix(X))
X_meta <- read.delim("data/microbiome_dataSet1_sampleInfo.tsv",row.names=1)
X_taxo <- read.delim("data/microbiome_dataSet1_taxonomy.tsv",row.names=1)
```

### Mark ambigous taxonomy as *incertae sedis*

To facilitate taxonomic-level analyses, we will update all ambiguous labels as incertae sedis.

```{r data_load_factors, iclude=FALSE}
#| echo: false
#| message: false
#| warning: false

#back up original taxonomy table
X_taxo_old <- X_taxo

#update all ambiguities in phylum, class, order, family and genus
X_taxo$phylum <- ifelse(grepl("/", X_taxo$phylum), "incertae sedis", X_taxo$phylum)
X_taxo$phylum <- ifelse(grepl("incertae", X_taxo$phylum, ignore.case = TRUE), 
                         "incertae sedis", 
                         X_taxo$phylum)
X_taxo$class <- ifelse(grepl("/", X_taxo$class), "incertae sedis", X_taxo$class)
X_taxo$class <- ifelse(grepl("incertae", X_taxo$class, ignore.case = TRUE), 
                         "incertae sedis", 
                         X_taxo$class)
X_taxo$order <- ifelse(grepl("/", X_taxo$order), "incertae sedis", X_taxo$order)
X_taxo$order <- ifelse(grepl("incertae", X_taxo$order, ignore.case = TRUE), 
                         "incertae sedis", 
                         X_taxo$order)
X_taxo$family <- ifelse(grepl("/", X_taxo$family), "incertae sedis", X_taxo$family)
X_taxo$family <- ifelse(grepl("incertae", X_taxo$family, ignore.case = TRUE), 
                         "incertae sedis", 
                         X_taxo$family)
X_taxo$genus <- ifelse(grepl("/", X_taxo$genus), "incertae sedis", X_taxo$genus)
X_taxo$genus <- ifelse(grepl("incertae", X_taxo$genus, ignore.case = TRUE), 
                         "incertae sedis", 
                         X_taxo$genus)

as.data.frame(table(X_taxo$phylum))
as.data.frame(table(X_taxo$class))
as.data.frame(table(X_taxo$order))
as.data.frame(table(X_taxo$family))
as.data.frame(table(X_taxo$genus))

```

### Basic Filtering

Find any null/NA values and set them to 0. Eliminate all-zero rows and columns. Check for any negative values.

```{r data_trns, include=FALSE}
#| echo: false
#| message: false
#| warning: false

#NAs to zeros
X[is.na(X)] <- 0

#remove all-zero samples
aux_allzero <- which(rowSums(X)==0)
if(length(aux_allzero)>0){
  X <- X[-aux_allzero,]
}

#remove all-zero taxa
aux_allzero <- which(colSums(X)==0)
if(length(aux_allzero)>0){
  X <- X[,-aux_allzero]
}

rm(aux_allzero)

if(any(X<0)){
  print("X has negative values. DO NOT CONTINUE.")
}

I <- dim(X)[1]; J <- dim(X)[2]
```

### Filtering rare taxa

Find and remove rare taxa: OTUs with only one count across all samples

```{r EDA_filter}
#| echo: false
#| message: false
#| warning: false

aux_rare <- which(colSums(X)==1)
X <- X[,-aux_rare]

#update dimension variables I and J
I <- dim(X)[1]; J <- dim(X)[2]

rm(aux_rare)
```

### Preprocessing

#### Binary data matrix

$$
f(x) =
\begin{cases} 
1 & \text{if } x > 0, \\
0 & \text{otherwise.}
\end{cases}
$$

```{r pre_bin}
#| echo: false
#| message: false
#| warning: false

Xbin <- X
Xbin[Xbin>0] <- 1

heatmap(Xbin, Rowv=NA, Colv=NA, scale="none", main="X (filtered + binarized)", cexRow=0.5, cexCol=0.1)
```

# Bibit

## Global objects

```{r pre_objects}
#| echo: false
#| message: false
#| warning: false

#list with all biclusters
biclusters <- vector("list")

#list for each bicluster, with 7 objects each
rows <- list()
cols <- list()
pattern <- ''

biclusters_sub <- list('rows'=rows,'cols'=cols,'pattern'=pattern)

stats_bc <- matrix(nrow=0,ncol=3)
aux_colnames <- c('bc','rows','cols')
colnames(stats_bc) <- aux_colnames

rm(aux_colnames)
```

## Functions

Encoding, searching and decoding will be called from a loop over the different bit word lengths, MNR and MNC

### General and formatting

Takes seconds as an input, returns minutes if \>= 60 seconds, hours if \>= 3600 seconds.

```{r func_format}
#| echo: false
#| message: false
#| warning: false


format_time <- function(seconds) {
  if (seconds < 60) {
    return(paste(round(seconds, 2), "seconds"))
  } else if (seconds < 3600) {
    minutes <- seconds / 60
    return(paste(round(minutes, 2), "minutes"))
  } else {
    hours <- seconds / 3600
    return(paste(round(hours, 2), "hours"))
  }
}

```

### Matrix: off-diagonal maximum

Used during merging. Takes a diagonal matrix of inter-cluster Jaccard indices and returns the highest value and the corresponding positions.

```{r func_general}
#| echo: false
#| message: false
#| warning: false

#check the highest off-diagonal value of a square matrix
off_diagonal_max <- function(matrix) {
  n <- nrow(matrix)
  aux_max <- -1
  matches <- list()
  
  for (i in 1:(n-1)) {
    for (j in (i+1):n) {
      if (i != j) {
        if (matrix[i, j] > aux_max) {
          aux_max <- matrix[i, j]
          matches <- list(list(row = i, col = j))
        } else if (matrix[i, j] == aux_max) {
          matches <- c(matches, list(list(row = i, col = j)))
        }
      }
    }
  }
  
  return(list(value = aux_max, positions = matches))
}


```

### Encoding

Phase 1 of Bibit. Takes a binarized matrix as input and returns a matrix encoded into bit words for faster computation when comparing row contents.

```{r func_encoding}
#| echo: false
#| message: false
#| warning: false

bb_encoding <- function(Xbin,I,J,bwl){
  time_enc_start <- Sys.time()

    #empty matrix to store the bitword-encoded matrix
  Xbw <- matrix(, nrow = 0, ncol = ceiling(J/bwl))
  
  for(i in 1:I){
    m <- 1
    aux_row <- c()
    for(j in seq(1,J,by=bwl)){
      aux_bitword <- ''
      for(k in 0:(bwl-1)){
        if(j+k <= J){
          aux_bitword <- paste0(aux_bitword,Xbin[i,j+k])
        }
      }
      aux_row[m] <- strtoi(aux_bitword, base=2)
      if(m<ceiling(J/bwl)){
        m <- m+1
      }
    }
    Xbw <- rbind(Xbw,aux_row)
  }
  
  time_enc_end <- Sys.time()
  time_enc <- round(as.numeric(difftime(time_enc_end, time_enc_start, units = "secs")),4)
  
  out_encoding <- vector("list",2)
  out_encoding[[1]] <- Xbw
  out_encoding[[2]] <- time_enc

  return(out_encoding)
}

```

### Searching

Phase 2 of Bibit. Takes the bitword-encoded matrix from the encoding phase, mnr and mnc. For every pair of rows (r~m~,r~n~), a pattern $\rho_{mn}$ is formed by performing a bitwise AND operation. If the number of 1s in $\rho_{mn}$ is \>= mnc, a bitwise AND operation is performed on each of the remainder rows of the encoded matrix. If the resulting pattern is equal to the original pattern, the row gets added to a temporary bicluster. After checking every remainder row, if the total number of rows in the temporary bicluster \>= mnr, the temporary bicluster is added to the global list of biclusters. The searching phase finishes when the pattern has been compared with all remainder rows.

Comments make reference to the line numbers of Algorithm 1

```{r func_searching}
#| echo: false
#| message: false
#| warning: false

bb_searching <- function(Xbw,mnr,mnc){

  time_sch_start <- Sys.time()

#empty object to store patterns
found_patterns <- c()    

# Generate all row pairs
aux_rowpairs <- combn(1:I, 2)

# ALGORITHM 1
# 1. for every rows pair (r_m,r_n) do
for(i in 1:ncol(aux_rowpairs)){
  m <- aux_rowpairs[1,i]
  n <- aux_rowpairs[2,i]
  #print(paste0('comparing r',m,' and r',n))
  
#2. rho_mn = r_m ^ r_n  
  rho_mn <- bitwAnd(Xbw[m,],Xbw[n,])

  # 2.1 get number of 1's in rho_mn:
  # 2.1.1 convert integer bitwords to binary strings of length bwl
  bw2bin <- sapply(rho_mn, function(x) {
    binary <- as.integer(intToBits(x))  #bitword: integer to binary (length 32)
    binary <- rev(binary[1:bwl])       #select the first bwl bits and reverse
    paste(binary, collapse = "")       #single bwl-bit long string
  })
  
  # 2.1.2 from binary strings to J-length pattern string
  n_bw <- length(bw2bin) #number of bitwords
  if(n_bw*bwl!=J){ #if number of bitwords*bwl != J
    nbits_tail <- bwl-(n_bw*bwl-J) #get number of necessary encoding bits at last bitword
    bw2bin[n_bw] <- str_sub(bw2bin[n_bw],-nbits_tail) #trim last bitword to encoding bits
  }
  pattern <- paste(bw2bin,collapse='') #binarized bitwords to J-length binary string

  # 2.2 (FOR LINE 3) get number of 1s from J-length pattern string
  pattern_1s <- str_count(pattern,'1')

  # 2.3 (FOR LINE 3) check if the pattern is new
  if ((pattern %in% found_patterns)){
      flag.new <- 0
      found_patterns <- c(found_patterns, pattern)
  }else{
    flag.new <- 1
  }
#3. if pattern is new and number of 1s in pattern >=mnc:
  #print(paste0('rho_',m,'_',n,': flag.new=',flag.new,' cols=',pattern_1s))
  if(flag.new & pattern_1s>=mnc){
#4. create and populate  Bic_mn bicluster
    biclusters_mn <- biclusters_sub #empty object to populate
    biclusters_mn[['rows']] <- c(m,n) #store rows M
    
    aux_chars <- strsplit(pattern, "")[[1]] #split pattern as vector to find 1s
    biclusters_mn[['cols']] <- which(aux_chars=='1') #store 1s as columns N
    
    biclusters_mn[['pattern']] <- rho_mn #store bitword pattern

    
    
    
   
}#5. for every remainder row q in Q
      Q <- 1:I
      Q <- setdiff(Q,biclusters_mn[['rows']])
      for(q in Q){
#6. if row_q AND pattern_mn == pattern_mn        
        rho_mnq <- bitwAnd(Xbw[q,],rho_mn)


        if(identical(rho_mnq,rho_mn)){
#7. add q to rows in Bic_mn          
          biclusters_mn[['rows']] <- c(biclusters_mn[['rows']],q)
        }
      }
#10. if n_rows in Bic_mn >= mnr, add to biclusters list      
      #print(paste0('rows=',length(biclusters_mn[['rows']])))
      if(length(biclusters_mn[['rows']])>=mnr){
          biclusters <- append(biclusters,list(biclusters_mn))
      }
  }


time_sch_end <- Sys.time()
  time_sch <- round(as.numeric(difftime(time_sch_end, time_sch_start, units = "secs")),4)
  
  out_searching <- vector("list",2)
  out_searching[[1]] <- biclusters
  out_searching[[2]] <- time_sch

  return(out_searching)
}
```

### Recursive Merging

Group of functions to recursively merge biclusters above a specified Jaccard index threshold over a specified measure (rows, columns, average). Returns a list of merged biclusters.

```{r merging_recursive_func}
#| echo: false
#| message: false
#| warning: false

#jaccard index between two biclusters
bc_jaccard <- function(bc_i,bc_j,measure){
  rows_I <- length(intersect(bc_i[['rows']],bc_j[['rows']]))
  rows_U <- length(union(bc_i[['rows']],bc_j[['rows']]))
  cols_I <- length(intersect(bc_i[['cols']],bc_j[['cols']]))
  cols_U <- length(union(bc_i[['cols']],bc_j[['cols']]))
  
  rows_jaccard <- rows_I/rows_U
  cols_jaccard <- cols_I/cols_U
  
  return(
    switch(measure,
    'rows'=rows_jaccard,
    'cols'=cols_jaccard,
    'avg'=mean(c(rows_jaccard,cols_jaccard)))
  )
  
  
}


#merge two biclusters
bc_merge <- function(bc_i,bc_j){
  list(
    rows = union(bc_i[['rows']],bc_j[['rows']]),
    cols = union(bc_i[['cols']],bc_j[['cols']]),
    initial_bc = sort(union(bc_i[['initial_bc']],bc_j[['initial_bc']]))
  )
}

#recursive merging
bc_merge_r <- function(biclusters,threshold,measure){
  n <- length(biclusters)
  if(n<=1){
    return(biclusters)
  }else{
    print(paste(n,'biclusters left...'))
    jaccard_matrix <- matrix(0,n,n)
    for(i in 1:(n-1)){
      for(j in (i+1):n){
        #print(paste0('i=',i,' j=',j))
        jaccard_matrix[i,j] <- bc_jaccard(biclusters[[i]],biclusters[[j]],measure)
      }
    }
    
    #find the maximum overlap
    jaccard_max <- max(jaccard_matrix)
    
    if(jaccard_max<=threshold){
      #if no pairs above threshold, return original bicluster list
      return(biclusters)
    }else{
      #find the pair of biclusters to merge
      max_indices <- which(jaccard_matrix == jaccard_max, arr.ind = TRUE)[1,]
      i <- max_indices[1]
      j <- max_indices[2]
      bc_i <- biclusters[[i]]
      bc_j <- biclusters[[j]]
      print(paste('merging biclusters',bc_i$initial_bc,'and',bc_j$initial_bc))
      #merge biclusters
      bc_merged <- bc_merge(bc_i,bc_j)
      
      #create new biclusters list
      biclusters_new <- biclusters[-c(i,j)]
      biclusters_new[[length(biclusters_new) + 1]] <- bc_merged
      
      #recursive call
      bc_merge_r(biclusters_new,threshold,measure)
    }
  }
}

```

## Set Global Parameters

After analyzing different run parameters, we choose a balance between:

-   Bit word length (bwl): longer words lead to shorter times. The maximum useful length is ceiling(log2(n))

-   Minimum number of rows (mnr): greater than 1 (a bicluster can be formed with a group at least two samples).

-   Minimum number of columns (mnc): greater than 1. A group of samples is allowed have at least 2 OTUs in common.

Across multiple combinations (see Bibit_supplement.qmd), we found that setting mnr=4 and mnc=10 with bwl=10 assigns all samples to a bicluster while managing to assign close to the maximum number of columns for non-minimal-sized parameters. bwl set to 10 showed the fastest performance

```{r global_params}
#| echo: false
#| message: false
#| warning: false

I = nrow(X)
J = ncol(X)

#bitword length
bwl <- (ceiling(log2(J)))
mnr <- 4
mnc <- 10

```

## 

## Main Loop

Runs bibit with the set parameters

```{r bibit_main}
time_global_start <- Sys.time()

print(paste0('running Bibit using bwl=',bwl,' mnr=',mnr,' mnc=',mnc))

out_encoding <- bb_encoding(Xbin,I,J,bwl)
Xbw <- out_encoding[[1]]
time_enc <- out_encoding[[2]]
print(paste0('      encoding finished in ',time_enc,' s'))
      
out_searching <- bb_searching(Xbw,mnr,mnc)
biclusters <- out_searching[[1]]
time_sch <- out_searching[[2]]
print(paste0('      searching finished in ',time_sch,' s'))

      n_biclust <- length(biclusters)
      if(n_biclust>0){
        print('analyzing found biclusters...')
  
        #get the rows and columns in the biclusters
        bc_allrows <- c()
        bc_allcols <- c()

        for(i in 1:n_biclust){
          bc_rows <- biclusters[[i]][["rows"]]
          bc_cols <- biclusters[[i]][["cols"]]
          bc_allrows <- append(bc_allrows,bc_rows)
          bc_allcols <- append(bc_allcols,bc_cols)
          bc_allrows <- unique(bc_allrows)
          bc_allcols <- unique(bc_allcols)
          aux_stats_bc_row <- c(i,length(bc_rows),length(bc_cols))
          stats_bc <- as.data.frame(rbind(stats_bc,aux_stats_bc_row))
        }

        n_rows <- length(unique(bc_allrows))
        n_cols <- length(unique(bc_allcols))
      }

for(i in 1:n_biclust){
  biclusters[[i]][['C']] <- X[biclusters[[i]][["rows"]],biclusters[[i]][["cols"]]]
}
      
time_global_end <- Sys.time()
time_global <- round(as.numeric(difftime(time_global_end, time_global_start, units = "secs")),4)

time_string <- format_time(time_global)

print(paste('found',n_rows,'rows and',n_cols,'columns across',n_biclust,'biclusters in', time_string))
#print(paste(m-n_rows,'rows and',n-n_cols,'columns are not part of any biclusters'))
print(paste(if(I-n_rows>0){paste(I-n_rows,'rows')}else{paste('')},
            if(J-n_cols>0){paste(J-n_cols,'columns')}else{paste('')},
            'are not part of any biclusters'))

rm(out_encoding,out_searching,time_enc,time_sch,time_global,time_global_start,time_global_end,time_string,i,aux_stats_bc_row)
```

## Cluster Analysis

### Descriptive statistics

Histograms:

-   Number of rows and columns per cluster.

-   Cluster dimensions (rows \* columns).

-   Cluster sparsity: proportion of zeros per cluster.

```{r bc_stats}
hist(stats_bc$rows,breaks=n_biclust,main="Number of rows per cluster")
hist(stats_bc$cols,breaks=n_biclust,main="Number of columns per cluster")

stats_bc$dim <- stats_bc$rows*stats_bc$cols
hist(stats_bc$dim,breaks=n_biclust,main="Cluster dimensions (row*col)")

stats_bc$sparsity <- 0
for(bcc in 1:length(biclusters)){
  stats_bc[bcc,'sparsity'] <- length(which(biclusters[[bcc]][["C"]]==0))/stats_bc[bcc,'dim']
}
hist(stats_bc$sparsity,breaks=n_biclust,main="Cluster sparsity")

```

### Compute inter-cluster Jaccard index

For all found clusters, compute a pairwise Jaccard index:

-   At the row level: IoU~rows~

-   At the column level: IoU~cols~

-   Average of both matrices

Output: symmetric matrices

```{r bc_iou}
#| echo: false
#| message: false
#| warning: false

# housekeeping: delete IoU matrices for reruns
biclusters[['iou_rows']] <- NULL
biclusters[['iou_cols']] <- NULL
biclusters[['iou_avg']] <- NULL
biclusters[['iou_rows_max']] <- NULL
biclusters[['iou_cols_max']] <- NULL
biclusters[['iou_avg_max']] <- NULL

# compute IoU matrices for rows and for columns
n_biclust <- length(biclusters)
if(n_biclust>1){
  print(paste0('computing Jaccard index between ',n_biclust,' biclusters...'))
  aux_intersect_rows <- matrix(0,nrow=n_biclust,ncol=n_biclust)
  aux_intersect_cols <- matrix(0,nrow=n_biclust,ncol=n_biclust)
  aux_union_rows <- matrix(0,nrow=n_biclust,ncol=n_biclust)
  aux_union_cols <- matrix(0,nrow=n_biclust,ncol=n_biclust)
  aux_iou_rows <- matrix(0,nrow=n_biclust,ncol=n_biclust)
  aux_iou_cols <- matrix(0,nrow=n_biclust,ncol=n_biclust)
  
  for(bc in 1:n_biclust){
    for(bcc in 1:n_biclust){
      aux_intersect_rows[bc,bcc] <- length(intersect(biclusters[[bc]][["rows"]],
                                                     biclusters[[bcc]][["rows"]]))
      aux_intersect_cols[bc,bcc] <- length(intersect(biclusters[[bc]][["cols"]],
                                                     biclusters[[bcc]][["cols"]]))
      aux_union_rows[bc,bcc] <- length(union(biclusters[[bc]][["rows"]],
                                             biclusters[[bcc]][["rows"]]))
      aux_union_cols[bc,bcc] <- length(union(biclusters[[bc]][["cols"]],
                                             biclusters[[bcc]][["cols"]]))
      aux_iou_rows[bc,bcc] <- round(aux_intersect_rows[bc,bcc]/aux_union_rows[bc,bcc],4)
      aux_iou_cols[bc,bcc] <- round(aux_intersect_cols[bc,bcc]/aux_union_cols[bc,bcc],4)
    }

    # store computed IoU matrices
    biclusters[["iou_rows"]] <- aux_iou_rows
    biclusters[["iou_cols"]] <- aux_iou_cols
    
    # get highest IoU: jaccard index and bicluster indices
    biclusters[["iou_rows_max"]] <- off_diagonal_max(aux_iou_rows)
    biclusters[["iou_cols_max"]] <- off_diagonal_max(aux_iou_cols)
    
  }
}

# average of both IoU matrices
aux_iou_avg <- matrix(0,nrow=n_biclust,ncol=n_biclust)
for(i in 1:n_biclust){
  for(j in 1:n_biclust){
    aux_iou_avg[i,j] <- (biclusters[['iou_rows']][i,j]+biclusters[['iou_cols']][i,j])/2
  }
}

# store computed average IoU matrix
biclusters[['iou_avg']] <- aux_iou_avg
biclusters[["iou_avg_max"]] <- off_diagonal_max(aux_iou_avg)

aux_pos_rows <- unlist(biclusters[["iou_rows_max"]][["positions"]])
aux_pos_cols <- unlist(biclusters[["iou_cols_max"]][["positions"]])
aux_pos_avg <- unlist(biclusters[["iou_avg_max"]][["positions"]])

print(paste0('highest row overlap: IoU=', biclusters[["iou_rows_max"]][["value"]]))
bc_i <- aux_pos_rows[1]
bc_j <- aux_pos_rows[2]

aux_intersect_rows <- intersect(biclusters[[aux_pos_rows[1]]][['rows']],biclusters[[aux_pos_rows[2]]][['rows']])
aux_intersect_cols <- intersect(biclusters[[aux_pos_rows[1]]][['cols']],biclusters[[aux_pos_rows[2]]][['cols']])

print(paste(paste('clusters',paste(aux_pos_rows, collapse=" & ")),'have',
            length(aux_intersect_rows),'samples &',
            length(aux_intersect_cols),'features in common'))

print('----------------------------------------')

print(paste0('highest column overlap: IoU=', biclusters[["iou_cols_max"]][["value"]]))
bc_i <- aux_pos_cols[1]
bc_j <- aux_pos_cols[2]

aux_intersect_rows <- intersect(biclusters[[aux_pos_cols[1]]][['rows']],biclusters[[aux_pos_cols[2]]][['rows']])
aux_intersect_cols <- intersect(biclusters[[aux_pos_cols[1]]][['cols']],biclusters[[aux_pos_cols[2]]][['cols']])

print(paste(paste('clusters',paste(aux_pos_cols, collapse=" & ")),'have',
            length(aux_intersect_rows),'samples &',
            length(aux_intersect_cols),'features in common'))

print('----------------------------------------')

print(paste0('highest averaged overlap: IoU=', biclusters[["iou_avg_max"]][["value"]]))
bc_i <- aux_pos_avg[1]
bc_j <- aux_pos_avg[2]

aux_intersect_rows <- intersect(biclusters[[aux_pos_avg[1]]][['rows']],biclusters[[aux_pos_avg[2]]][['rows']])
aux_intersect_cols <- intersect(biclusters[[aux_pos_avg[1]]][['cols']],biclusters[[aux_pos_avg[2]]][['cols']])

print(paste(paste('clusters',paste(aux_pos_avg, collapse=" & ")),'have',
            length(aux_intersect_rows),'samples &',
            length(aux_intersect_cols),'features in common'))

rm(aux_pos_rows,aux_pos_cols,aux_pos_avg,bc_i,bc_j,aux_intersect_rows,aux_intersect_cols,aux_union_rows,aux_union_cols,aux_iou_rows,aux_iou_cols,aux_iou_avg,bc,bcc)
rm(i,j)
```

#### Visualize IoU matrices as heatmaps

```{r heatmaps_iou}
#| echo: false
#| message: false
#| warning: false

heatmap(biclusters[["iou_rows"]], Rowv=NA, Colv=NA, scale="none", main='IoU: rows',cexRow=0.5,cexCol=0.5)
heatmap(biclusters[["iou_cols"]], Rowv=NA, Colv=NA, scale="none", main='IoU: columns',cexRow=0.5,cexCol=0.5)
heatmap(biclusters[["iou_avg"]], Rowv=NA, Colv=NA, scale="none", main='averaged IoU: rows&columns',cexRow=0.5,cexCol=0.52)

```

Individually check the resulting IoUs between any two biclusters

```{r IoU verify}
#| echo: false
#| message: false
#| warning: false

# edit to compare two biclusters:
bc_m <- 13
bc_n <- 16

# get number of rows and columns of both biclusters
bc_m_nrow <- length(biclusters[[bc_m]][['rows']])
bc_m_ncol <- length(biclusters[[bc_m]][['cols']])
bc_n_nrow <- length(biclusters[[bc_n]][['rows']])
bc_n_ncol <- length(biclusters[[bc_n]][['cols']])

# compute IoU_rows
aux_in_rows <- length(intersect(biclusters[[bc_m]][['rows']],biclusters[[bc_n]][['rows']]))
aux_un_rows <- length(union(biclusters[[bc_m]][['rows']],biclusters[[bc_n]][['rows']]))
aux_iou_rows <- aux_in_rows/aux_un_rows

# compute IoU_cols
aux_in_cols <- length(intersect(biclusters[[bc_m]][['cols']],biclusters[[bc_n]][['cols']]))
aux_un_cols <- length(union(biclusters[[bc_m]][['cols']],biclusters[[bc_n]][['cols']]))
aux_iou_cols <- aux_in_cols/aux_un_cols

# print result
print(paste0('biclusters ',bc_m,' (',bc_m_nrow,'x',bc_m_ncol,') and ',bc_n,' (',bc_n_nrow,'x',bc_n_ncol,') share ',
             aux_in_rows,' rows and ',aux_in_cols,' columns'))
print(paste0('IoU_rows=',round(aux_iou_rows,2),' | IoU_columns=',round(aux_iou_cols,2)))
```

### Dendrograms: visualize merging candidates

Converting the Jaccard indices into distances, display dendrograms shown possible merging candidates. Set a distance as possible merging threshold and draw it as a horizontal line to evaluate. The choice for linkage method and distance threshold is up to the researcher and findings in the particular data set and the desired "cut point" after viewing the dendrograms.

```{r dendro}
#| echo: false
#| message: false
#| warning: false

#Choose linkage method between: average, median, centroid, single, complete
aux_linkage <- 'average'

#set possible merging threshold
aux_thresh <- 0.7


# Jaccard index to distance (1 - Jaccard index)
distance_rows_m <- 1 - biclusters[['iou_rows']]
distance_cols_m <- 1 - biclusters[['iou_cols']]
distance_avg_m <- 1 - biclusters[['iou_avg']]

# hierarchical clustering
# convert distance matrix to "dist" vector
distance_rows_v <- as.dist(distance_rows_m)
distance_cols_v <- as.dist(distance_cols_m)
distance_avg_v <- as.dist(distance_avg_m)

# use average linkage
hc_rows <- hclust(distance_rows_v, method=aux_linkage)
hc_cols <- hclust(distance_cols_v, method=aux_linkage)
hc_avg <- hclust(distance_avg_v, method=aux_linkage)

# plot dendrograms
plot(hc_rows, labels = paste("BC", 1:n_biclust, sep=""),
     main=paste("Dendrogram: row-wise similarity",'\n',aux_linkage,'linkage'),
     xlab="Biclusters", ylab="Distance = 1 - Jrows(BCm,BCn)",sub='')
abline(h = aux_thresh, col = "red", lty = 2, lwd = 2)

plot(hc_cols, labels = paste("BC", 1:n_biclust, sep=""),
     main=paste("Dendrogram: column-wise similarity",'\n',aux_linkage,'linkage'),
     xlab="Biclusters", ylab="Distance = 1 - Jcols(BCm,BCn)",sub='')
abline(h = aux_thresh, col = "red", lty = 2, lwd = 2)

plot(hc_avg, labels = paste("BC", 1:n_biclust, sep=""),
     main=paste("Dendrogram: averaged row-column similarity",'\n',aux_linkage,'linkage'),
     xlab="Biclusters", ylab="Distance = 1 - Javerage(BCm,BCn)",sub='')
abline(h = aux_thresh, col = "red", lty = 2, lwd = 2)

```

Looking at the three plots, we can make a choice between merging at a distance threshold on the rows, columns or the average, as well as adjust the threshold to a desired level.

```{r dendro_rm}
#| echo: false
#| message: false
#| warning: false

rm(distance_rows_m,distance_rows_v,distance_cols_m,distance_cols_v,distance_avg_m,distance_avg_v)
```

### Run merging

Looking at the dendrograms, set the threshold and choose the measure dimension between "rows", "cols" and "avg".

```{r merging_recursive}

#set or update the threshold (look at dendrogram above) and set the measure
aux_thresh <- 0.7


#choose between rows, cols, avg
aux_measure <- 'cols'

if (!(aux_measure %in% c('rows','cols','avg'))) {
  print(paste('INVALID MEASURE SELECTED:',aux_measure,'// instead, use: rows, cols, avg'))
}else{
  # housekeeping: delete IoU matrices for reruns, empty stats_bc_merged
  biclusters[['iou_rows']] <- NULL
  biclusters[['iou_cols']] <- NULL
  biclusters[['iou_avg']] <- NULL
  biclusters[['iou_rows_max']] <- NULL
  biclusters[['iou_cols_max']] <- NULL
  biclusters[['iou_avg_max']] <- NULL
  
  stats_bc_merged <- matrix(nrow=0,ncol=4)
  aux_colnames <- c('bc','rows','cols')
  colnames(stats_bc_merged) <- c(aux_colnames,'n_mergedbc')
  
  
  time_merge_start <- Sys.time()
  
  if(n_biclust>0){
        print('analyzing found biclusters...')
  
        #get the rows and columns in the biclusters
        bc_allrows <- c()
        bc_allcols <- c()

        for(i in 1:n_biclust){
          bc_rows <- biclusters[[i]][["rows"]]
          bc_cols <- biclusters[[i]][["cols"]]
          bc_allrows <- append(bc_allrows,bc_rows)
          bc_allcols <- append(bc_allcols,bc_cols)
          bc_allrows <- unique(bc_allrows)
          bc_allcols <- unique(bc_allcols)
        }

        n_rows <- length(unique(bc_allrows))
        n_cols <- length(unique(bc_allcols))
      }
  
  #store the original bicluster numbering
  for(bc in 1:n_biclust){
    biclusters[[bc]][['initial_bc']] <- bc
  }
  
  print(paste0('Merging biclusters along Jaccard_',aux_measure,' with threshold=',aux_thresh,'...'))
  biclusters_merged <- bc_merge_r(biclusters, threshold = (1-aux_thresh), measure=aux_measure)
  n_biclust_merged <- length(biclusters_merged)
  
  time_merge_end <- Sys.time()
  time_merge <- round(as.numeric(difftime(time_merge_end, time_merge_start, units = "secs")),4)
  time_string <- format_time(time_merge)
  
  print(paste('merged',n_biclust,'biclusters into',n_biclust_merged,'in', time_string))
  
  for(i in 1:n_biclust_merged){
    bc_rows <- biclusters_merged[[i]][["rows"]]
    bc_cols <- biclusters_merged[[i]][["cols"]]
    bc_n_merged <- unique(biclusters_merged[[i]][["initial_bc"]])
    bc_allrows <- append(bc_allrows,bc_rows)
    bc_allcols <- append(bc_allcols,bc_cols)
    bc_allrows <- unique(bc_allrows)
    bc_allcols <- unique(bc_allcols)
    aux_stats_bc_row <- c(i,length(bc_rows),length(bc_cols),length(bc_n_merged))
    stats_bc_merged <- as.data.frame(rbind(stats_bc_merged,aux_stats_bc_row))
    }
}
rownames(stats_bc_merged) <- 1:n_biclust_merged

#update C matrix (subset of Xsqrt)
for(bcc in 1:n_biclust_merged){
  aux_rows <- biclusters_merged[[bcc]][['rows']]
  aux_cols <- biclusters_merged[[bcc]][['cols']]
  biclusters_merged[[bcc]][['C']] <- as.data.frame(X[aux_rows,aux_cols])
}

rm(time_merge_start,time_merge_end,time_merge,time_string)
```

```{r bc_merged_table}
#| echo: false
#| message: false
#| warning: false

stats_bc_merged$dim <- stats_bc_merged$rows*stats_bc_merged$cols
stats_bc_merged$sparsity <- 0
for(bcc in 1:length(biclusters_merged)){
  stats_bc_merged[bcc,'sparsity'] <- length(which(biclusters_merged[[bcc]][["C"]]==0))/stats_bc_merged[bcc,'dim']
}


#stats_bc_merged_table <- stats_bc_merged[,c('bc','rows','cols')]
#stats_bc_merged_table$density <- 1 - stats_bc_merged$sparsity

#stats_bc_merged_table
```

### Jaccard Index on merged clusters

Verify whether the merged biclusters maintain any significant overlaps.

```{r bc_iou_merged}
#| echo: false
#| message: false
#| warning: false

# housekeeping: delete IoU matrices for reruns
biclusters_merged[['iou_rows']] <- NULL
biclusters_merged[['iou_cols']] <- NULL
biclusters_merged[['iou_avg']] <- NULL
biclusters_merged[['iou_rows_max']] <- NULL
biclusters_merged[['iou_cols_max']] <- NULL
biclusters_merged[['iou_avg_max']] <- NULL

# compute IoU matrices for rows and for columns
n_biclust <- length(biclusters_merged)
if(n_biclust>1){
  print(paste0('computing Jaccard index between ',n_biclust,' biclusters...'))
  aux_intersect_rows <- matrix(0,nrow=n_biclust,ncol=n_biclust)
  aux_intersect_cols <- matrix(0,nrow=n_biclust,ncol=n_biclust)
  aux_union_rows <- matrix(0,nrow=n_biclust,ncol=n_biclust)
  aux_union_cols <- matrix(0,nrow=n_biclust,ncol=n_biclust)
  aux_iou_rows <- matrix(0,nrow=n_biclust,ncol=n_biclust)
  aux_iou_cols <- matrix(0,nrow=n_biclust,ncol=n_biclust)
  
  for(bc in 1:n_biclust){
    for(bcc in 1:n_biclust){
      aux_intersect_rows[bc,bcc] <- length(intersect(biclusters_merged[[bc]][["rows"]],
                                                     biclusters_merged[[bcc]][["rows"]]))
      aux_intersect_cols[bc,bcc] <- length(intersect(biclusters_merged[[bc]][["cols"]],
                                                     biclusters_merged[[bcc]][["cols"]]))
      aux_union_rows[bc,bcc] <- length(union(biclusters_merged[[bc]][["rows"]],
                                             biclusters_merged[[bcc]][["rows"]]))
      aux_union_cols[bc,bcc] <- length(union(biclusters_merged[[bc]][["cols"]],
                                             biclusters_merged[[bcc]][["cols"]]))
      aux_iou_rows[bc,bcc] <- round(aux_intersect_rows[bc,bcc]/aux_union_rows[bc,bcc],4)
      aux_iou_cols[bc,bcc] <- round(aux_intersect_cols[bc,bcc]/aux_union_cols[bc,bcc],4)
    }

    # store computed IoU matrices
    biclusters_merged[["iou_rows"]] <- aux_iou_rows
    biclusters_merged[["iou_cols"]] <- aux_iou_cols
    
    # get highest IoU: jaccard index and bicluster indices
    biclusters_merged[["iou_rows_max"]] <- off_diagonal_max(aux_iou_rows)
    biclusters_merged[["iou_cols_max"]] <- off_diagonal_max(aux_iou_cols)
    
  }
}

# average of both IoU matrices
aux_iou_avg <- matrix(0,nrow=n_biclust,ncol=n_biclust)
for(i in 1:n_biclust){
  for(j in 1:n_biclust){
    aux_iou_avg[i,j] <- (biclusters_merged[['iou_rows']][i,j]+biclusters_merged[['iou_cols']][i,j])/2
  }
}

# store computed average IoU matrix
biclusters_merged[['iou_avg']] <- aux_iou_avg
biclusters_merged[["iou_avg_max"]] <- off_diagonal_max(aux_iou_avg)

aux_pos_rows <- unlist(biclusters_merged[["iou_rows_max"]][["positions"]])
aux_pos_cols <- unlist(biclusters_merged[["iou_cols_max"]][["positions"]])
aux_pos_avg <- unlist(biclusters_merged[["iou_avg_max"]][["positions"]])

print(paste0('highest row overlap: IoU=', biclusters_merged[["iou_rows_max"]][["value"]]))
bc_i <- aux_pos_rows[1]
bc_j <- aux_pos_rows[2]

aux_intersect_rows <- intersect(biclusters_merged[[aux_pos_rows[1]]][['rows']],biclusters_merged[[aux_pos_rows[2]]][['rows']])
aux_intersect_cols <- intersect(biclusters_merged[[aux_pos_rows[1]]][['cols']],biclusters_merged[[aux_pos_rows[2]]][['cols']])

print(paste(paste('clusters',paste(aux_pos_rows, collapse=" & ")),'have',
            length(aux_intersect_rows),'samples &',
            length(aux_intersect_cols),'features in common'))

print('----------------------------------------')

print(paste0('highest column overlap: IoU=', biclusters_merged[["iou_cols_max"]][["value"]]))
bc_i <- aux_pos_cols[1]
bc_j <- aux_pos_cols[2]

aux_intersect_rows <- intersect(biclusters_merged[[aux_pos_cols[1]]][['rows']],biclusters_merged[[aux_pos_cols[2]]][['rows']])
aux_intersect_cols <- intersect(biclusters_merged[[aux_pos_cols[1]]][['cols']],biclusters_merged[[aux_pos_cols[2]]][['cols']])

print(paste(paste('clusters',paste(aux_pos_cols, collapse=" & ")),'have',
            length(aux_intersect_rows),'samples &',
            length(aux_intersect_cols),'features in common'))

print('----------------------------------------')

print(paste0('highest averaged overlap: IoU=', biclusters_merged[["iou_avg_max"]][["value"]]))
bc_i <- aux_pos_avg[1]
bc_j <- aux_pos_avg[2]

aux_intersect_rows <- intersect(biclusters_merged[[aux_pos_avg[1]]][['rows']],biclusters_merged[[aux_pos_avg[2]]][['rows']])
aux_intersect_cols <- intersect(biclusters_merged[[aux_pos_avg[1]]][['cols']],biclusters_merged[[aux_pos_avg[2]]][['cols']])

print(paste(paste('clusters',paste(aux_pos_avg, collapse=" & ")),'have',
            length(aux_intersect_rows),'samples &',
            length(aux_intersect_cols),'features in common'))

rm(aux_pos_rows,aux_pos_cols,aux_pos_avg,bc_i,bc_j,aux_intersect_rows,aux_intersect_cols,aux_union_rows,aux_union_cols,aux_iou_rows,aux_iou_cols,aux_iou_avg,bc,bcc)
rm(i,j)
```

#### Visualize IoU matrices as heatmaps

```{r heatmaps_iou}
#| echo: false
#| message: false
#| warning: false

heatmap(biclusters_merged[["iou_rows"]], Rowv=NA, Colv=NA, scale="none", main='IoU: rows')
heatmap(biclusters_merged[["iou_cols"]], Rowv=NA, Colv=NA, scale="none", main='IoU: columns')
heatmap(biclusters_merged[["iou_avg"]], Rowv=NA, Colv=NA, scale="none", main='averaged IoU: rows&columns')

```

#### Dendrograms: merged clusters (new numeration)

```{r dendro_merged}
#| echo: false
#| message: false
#| warning: false

# Jaccard index to distance (1 - Jaccard index)
distance_rows_m <- 1 - biclusters_merged[['iou_rows']]
distance_cols_m <- 1 - biclusters_merged[['iou_cols']]
distance_avg_m <- 1 - biclusters_merged[['iou_avg']]

# hierarchical clustering
# convert distance matrix to "dist" vector
distance_rows_v <- as.dist(distance_rows_m)
distance_cols_v <- as.dist(distance_cols_m)
distance_avg_v <- as.dist(distance_avg_m)

# use average linkage
hc_rows <- hclust(distance_rows_v, method="average")
hc_cols <- hclust(distance_cols_v, method="average")
hc_avg <- hclust(distance_avg_v, method="average")

#change merging threshold?
#aux_thresh <- 0.55

# plot dendrograms
plot(hc_rows, labels = paste("BC", 1:n_biclust, sep=""),
     main="Dendrogram: row-wise similarity",
     xlab="Biclusters", ylab="Distance = 1 - Jrows(BCm,BCn)",sub='')
abline(h = aux_thresh, col = "red", lty = 2, lwd = 2)

plot(hc_cols, labels = paste("BC", 1:n_biclust, sep=""),
     main="Dendrogram: column-wise similarity",
     xlab="Biclusters", ylab="Distance = 1 - Jcols(BCm,BCn)",sub='')
abline(h = aux_thresh, col = "red", lty = 2, lwd = 2)

plot(hc_avg, labels = paste("BC", 1:n_biclust, sep=""),
     main="Dendrogram: averaged row-column similarity",
     xlab="Biclusters", ylab="Distance = 1 - Javerage(BCm,BCn)",sub='')
abline(h = aux_thresh, col = "red", lty = 2, lwd = 2)

```

### Visualize all clusters in a matrix

#### Order rows and columns by bicluster

Get (reverse) order from the post-merge hierarchical clustering of biclusters

```{r X_bc_order}
#| echo: false
#| message: false
#| warning: false

#choose normal or reverse order
#aux_bc_order <- rev(hc_avg[["order"]])
aux_bc_order <- hc_avg[["order"]]

```

Compute union of rows and columns of biclusters following the chosen order

```{r X_bc_rowscols}
#| echo: false
#| message: false
#| warning: false

joinrows <- c()
joincols <- c()

for(bcc in aux_bc_order){
  aux_pos <- which(aux_bc_order==bcc)
  if(aux_pos<length(aux_bc_order)){
    aux_x <- aux_bc_order[aux_pos]
    aux_y <- aux_bc_order[aux_pos+1]
    if(aux_pos==1){
      #print(paste('comparing',aux_x,aux_y))
      
      #union without reordering. intersection first.
      aux_x_rows <- biclusters_merged[[aux_x]][['rows']]
      aux_x_cols <- biclusters_merged[[aux_x]][['cols']]
      
    }else{
      #print(paste('comparing previous merge with',aux_y))
      aux_x_rows <- aux_union_rows
      aux_x_cols <- aux_union_cols
      }
    aux_y_rows <- biclusters_merged[[aux_y]][['rows']]
    aux_y_cols <- biclusters_merged[[aux_y]][['cols']]
    
    aux_intersect_rows <- intersect(aux_x_rows,aux_y_rows)
    aux_setdiff_rows <- setdiff(union(aux_x_rows,aux_y_rows),aux_intersect_rows)
    aux_union_rows <- c(aux_intersect_rows,aux_setdiff_rows)

        aux_intersect_cols <- intersect(aux_x_cols,aux_y_cols)
    aux_setdiff_cols <- setdiff(union(aux_x_cols,aux_y_cols),aux_intersect_cols)
    aux_union_cols <- c(aux_intersect_cols,aux_setdiff_cols)

  }
}



rm(aux_pos,aux_x,aux_y,aux_x_rows,aux_x_cols,aux_y_rows,aux_y_cols,aux_intersect_rows,aux_intersect_cols,aux_setdiff_rows,aux_setdiff_cols)
```

#### Color-code elements by bicluster membership

Assigns the bicluster number to each element, assigns n_biclust + 1 to overlapping elements. Any unassigned element is set to zero.

```{r X_color_assign}
#| echo: false
#| message: false
#| warning: false

Xcolor <- X
Xcolor[,] <- NA
for(bcc in aux_bc_order){
  aux_rows <- biclusters_merged[[bcc]][['rows']]
  aux_cols <- biclusters_merged[[bcc]][['cols']]
  for(i in aux_rows){
    for(j in aux_cols){
      if(is.na(Xcolor[i,j])){ #if the element is empty, assign a cluster number as label
        Xcolor[i,j]<-bcc
      }else{ #if it already contains a value (overlap), assign a very large number as label
        Xcolor[i,j]<-n_biclust_merged+1
      }
      
    }
  }
}

Xcolor[is.na(Xcolor)] <- 0
```

#### Biclusters in matrix: clustered rows and columns only

Rows and columns ordered and grouped by cluster membership. Creates a rainbow pallette using the number of biclusters. Unassigned elements are black, overlapping elements are white.

```{r X_bc_sub}
#| echo: false
#| message: false
#| warning: false

aux_colors <- c('#000000',rainbow(n_biclust_merged),'#ffffff')

heatmap(Xcolor[aux_union_rows,aux_union_cols], Rowv=NA, Colv=NA, scale="none", main='Clusters in a subset of X',col = aux_colors,cexRow=0.5,cexCol=0.1)

colorLegend <- aux_colors
legend("topright",legend = c('none',1:n_biclust_merged,'overlap'),
       fill=colorLegend, title = "bicluster")


heatmap(Xbin[aux_union_rows,aux_union_cols], Rowv=NA, Colv=NA, scale="none", main='Binarized X (clustered columns)',cexRow=0.5,cexCol=0.1)

```

#### Biclusters in X

Rows and columns ordered and grouped by cluster membership, unclustered columns on the left, unclustered rows on the bottom (if any)

```{r X_bc}
#| echo: false
#| message: false
#| warning: false

aux_union_rows_X <- c(aux_union_rows,setdiff(1:I,aux_union_rows))
aux_union_cols_X <- c(aux_union_cols,setdiff(1:J,aux_union_cols))

heatmap(Xcolor[aux_union_rows_X,aux_union_cols_X], Rowv=NA, Colv=NA, scale="none", main='Clusters in X',col = aux_colors,cexRow=0.5,cexCol=0.1)

colorLegend <- aux_colors
legend("topright",legend = c('none',1:n_biclust_merged,'overlap'),
       fill=colorLegend, title = "bicluster")
```

### Visualize individual clusters

```{r bc_all_view}
#| echo: false
#| message: false
#| warning: false

for(bcc in 1:n_biclust){
  
  if(stats_bc_merged[stats_bc_merged$bc==bcc,'rows']<=20){aux_cexrow<-1}else{aux_cexrow<-0.5}
  if(stats_bc_merged[stats_bc_merged$bc==bcc,'cols']<=40){aux_cexcol<-1}else{
    if(stats_bc_merged[stats_bc_merged$bc==bcc,'cols']<=100){
      aux_cexcol<-0.25
      }else{
        aux_cexcol<-0.1
      }
    }
  
  aux_matrix <- as.matrix(sqrt(sqrt(biclusters_merged[[bcc]][['C']])))
  heatmap(aux_matrix, Rowv=NA, Colv=NA, scale="none", main=paste('Cluster',bcc),cexRow=aux_cexrow,cexCol=aux_cexcol)
}


```

### Biological relevance

Identify by index: study groups and dysbiotic status in the samples, class and family of taxa.

Sample subsets:

-   3 study groups

-   2 dysbiotic status

Taxonomy:

-   10 phyla

-   17 classes

-   48 families

```{r bio_samples}
#| echo: false
#| message: false
#| warning: false

for(bcc in 1:n_biclust){
    aux_meta <- X_meta[biclusters_merged[[bcc]][["rows"]],]
    biclusters_merged[[bcc]][['diagnosis']] <- as.matrix(table(aux_meta$diagnosis))
    biclusters_merged[[bcc]][['dysbiotic']] <- as.matrix(table(aux_meta$dysbiotic))
}

ind_meta_diag <- vector('list')
for(d in unique(X_meta$diagnosis)){
  ind_meta_diag[[d]] <- which(X_meta$diagnosis==d)
}

X_meta$dysbiotic <- ifelse(X_meta$dysbiotic,'T','F')
ind_meta_dysb <- vector('list')
for(d in unique(X_meta$dysbiotic)){
  ind_meta_dysb[[d]] <- which(X_meta$dysbiotic==d)
}
```

```{r bio_taxa}
#| echo: false
#| message: false
#| warning: false

for(bcc in 1:n_biclust){
    aux_taxo <- X_taxo[biclusters_merged[[bcc]][["cols"]],]
    biclusters_merged[[bcc]][['phylum']] <- as.matrix(table(aux_taxo$phylum))
    biclusters_merged[[bcc]][['class']] <- as.matrix(table(aux_taxo$class))
    biclusters_merged[[bcc]][['family']] <- as.matrix(table(aux_taxo$family))
}

ind_taxo_phylum <- vector('list')
for(t in unique(X_taxo$phylum)){
  ind_taxo_phylum[[t]] <- which(X_taxo$phylum==t)
}
ind_taxo_class <- vector('list')
for(t in unique(X_taxo$class)){
  ind_taxo_class[[t]] <- which(X_taxo$class==t)
}
ind_taxo_family <- vector('list')
for(t in unique(X_taxo$family)){
  ind_taxo_family[[t]] <- which(X_taxo$family==t)
}
```

For each bicluster, list the distribution of diagnosis and dysbiosis.

As counts:

```{r bc_diag_dysb}
#| echo: false
#| message: false
#| warning: false


for(bcc in 1:n_biclust){
  print(paste0('BC',bcc))
  print('diagnosis:')
  for(d in unique(X_meta$diagnosis)){
    aux_intersect<- length(intersect(biclusters_merged[[bcc]][['rows']],ind_meta_diag[[d]]))
    if(aux_intersect>0){print(paste(d,':',aux_intersect))}
  }
  print('dysbiotic:')
  for(d in unique(X_meta$dysbiotic)){
    aux_intersect <- length(intersect(biclusters_merged[[bcc]][['rows']],ind_meta_dysb[[d]]))
    if(aux_intersect>0){print(paste(d,':',aux_intersect))}
  }
  print('--------------------')
}
```

Identifying individual samples:

```{r bc_diag_dysb_table}
for(bcc in 1:n_biclust){
  aux_rows <- biclusters_merged[[bcc]][['rows']]
  print(X_meta[aux_rows,c('diagnosis','dysbiotic')])
}
```

For each bicluster, list the distribution by phylum, class, family.

```{r bc_taxo}
for(bcc in 1:n_biclust){
  print(paste0('BC',bcc))
  print('phylum:')
  for(d in unique(X_taxo$phylum)){
    aux_intersect<- length(intersect(biclusters_merged[[bcc]][['cols']],ind_taxo_phylum[[d]]))
    if(aux_intersect>0){print(paste(d,':',aux_intersect))}
  }
  print('--------------------')
  print('class:')
  for(d in unique(X_taxo$class)){
    aux_intersect<- length(intersect(biclusters_merged[[bcc]][['cols']],ind_taxo_class[[d]]))
    if(aux_intersect>0){print(paste(d,':',aux_intersect))}
  }
  print('--------------------')
  print('family:')
  for(d in unique(X_taxo$family)){
    aux_intersect<- length(intersect(biclusters_merged[[bcc]][['cols']],ind_taxo_family[[d]]))
    if(aux_intersect>0){print(paste(d,':',aux_intersect))}
  }
  print('----------------------------------------')
}
```

### 

### Inspect a single cluster

Create a temporary object with the bicluster, selected by number.

```{r bc_singgle}

BC <- 1
aux_bc <- as.matrix(biclusters_merged[[BC]][['C']])

```

```{r hc_inspect_meta}
#| echo: false
#| message: false
#| warning: false

#input a study group and a dysbiotic status of interest
aux_diag <- 'IBD_CD'
aux_dysb <- 'T'

#number of results for median taxonomic abundance
aux_n <- 10

#convert TRUE and FALSE to T and F
X_meta$dysbiotic <- ifelse(X_meta$dysbiotic,'T','F')

#get row names of all samples and of nonzero taxa in the cluster
aux_samples <- rownames(aux_bc)
aux_taxa <- colnames(aux_bc)

#get the metadata and taxonomy of the clustered rows and the nonzero taxa
X_meta_bc <- X_meta[aux_samples,]
X_taxo_bc <- X_taxo[aux_taxa,]

#visualize the group distribution of the selection of samples
as.data.frame(table(X_meta_bc$diagnosis)) #samples by diagnosis
as.data.frame(table(X_meta_bc$dysbiotic)) #samples by dysbiotic status
as.data.frame(table(X_taxo_bc$class)) #taxa by class
as.data.frame(table(X_taxo_bc$family)) #taxa by family

#select a diagnosis and get a submatrix from the bicluster
X_meta_bc_diag <- X_meta_bc[X_meta_bc$diagnosis==aux_diag,]
aux_bc_diag <- aux_bc[rownames(X_meta_bc_diag),]

#select by dysbiotic status and get a submatrix from the bicluster
X_meta_bc_dysb <- X_meta_bc[X_meta_bc$dysbiotic==aux_dysb,]
aux_bc_dysb <- aux_bc[rownames(X_meta_bc_dysb),]

#visualize the selections to find patterns
heatmap(aux_bc_diag, Rowv=NA, Colv=NA, scale="none", main=paste0('Cluster',BC,' (',aux_diag,')'),cexRow=1,cexCol=0.25)

heatmap(aux_bc_dysb, Rowv=NA, Colv=NA, scale="none", main=paste0('Cluster',BC, '(',
                                                                ifelse(aux_dysb=='T','','non-'),'dysbiotic)'),cexRow=1,cexCol=0.25)

#keep only nonzero taxa
aux_bc_diag <- aux_bc_diag[,aux_taxa]
aux_bc_dysb <- aux_bc_dysb[,aux_taxa]

#get the median abundance of taxa by diagnosis and by dysbiotic status
#by diagnostic
aux_bc_diag_colsum_median <- apply(aux_bc_diag,2,median)
print(paste('Top ',aux_n,' median abundance taxa in',paste0('BC',BC),aux_diag,':'))
head(sort(aux_bc_diag_colsum_median[aux_bc_diag_colsum_median>0], decreasing = TRUE),n=aux_n)
print('----------------------------------------')
print(paste('Bottom ',aux_n,' median abundance taxa in',paste0('BC',BC),aux_diag,':'))
head(sort(aux_bc_diag_colsum_median[aux_bc_diag_colsum_median>0], decreasing = FALSE),n=aux_n)

print('========================================')
#by dysbiosis
aux_bc_dysb_colsum_median <- apply(aux_bc_dysb,2,median)
print(paste0('Top ',aux_n,' median abundance taxa in ',paste0('BC',BC),ifelse(aux_dysb=='T','','non-'),' dysbiotic:'))
head(sort(aux_bc_dysb_colsum_median[aux_bc_dysb_colsum_median>0], decreasing = TRUE),n=aux_n)
print('----------------------------------------')
print(paste0('Bottom ',aux_n,' median abundance taxa in ',paste0('BC',BC),ifelse(aux_dysb=='T','','non-'),'d ysbiotic:'))
head(sort(aux_bc_dysb_colsum_median[aux_bc_dysb_colsum_median>0], decreasing = FALSE),n=aux_n)

```

### 
