---
title: "Microbiome Data Analysis Using Biclustering Approaches - Initial Exploratory Data Analysis"
author: "Italo Rossi del Aguila"
format: pdf
editor: visual
---

```{r load_pk, include=FALSE}
#| echo: false
#| message: false
#| warning: false
library(tidyr)
library(dplyr)
library(vegan) #for rarefaction
library(cluster)
```

# Initial Exploratory Analysis

## 16S IBD

The present work will use a data set derived from Lloyd-Price et al. (2019). This study investigates the gut microbiome of individuals with two types of inflammatory bowel disease, namely Crohn's disease (IBD_CD) and ulcerative colitis (IBD_UC) in comparison to healthy individuals (HC, for healthy controls). Specifically, we will be using only one sample per individual for all three study groups: 43 samples and 668 OTUs.

The chosen data set has been rarefied (downsampled) so all samples have the same sequencing depth: 1538 reads. Three study groups are identified in the accompanying metadata: two groups of inflammatory bowel disease groups: Crohn's disease (IBD_CD -- 17 samples), ulcerative colitis (IBD_UC -- 12 samples) and healthy controls (HC, 14 samples). Each sample receives a dysbiosis score, from which a dysbiotic status is determined as TRUE or FALSE: this is calculated by taking the median Bray-Curtis dissimilarity of each sample compared to a reference sample set (excluding samples taken from the same subject; details in the Methods section of Lloyd-Price et al., 2019).

### Import data

-   **X**: nonnegative integer data matrix

    -   Transpose so samples are in the rows and features in the columns

-   **X~meta~**: sample metadata

-   **X~taxo~**: feature taxonomy

```{r data_load, include=FALSE}
#| echo: false
#| message: false
#| warning: false
X <- read.delim("data/microbiome_dataSet1.tsv",row.names=1)
X <- t(as.matrix(X))
X_meta <- read.delim("data/microbiome_dataSet1_sampleInfo.tsv",row.names=1)
X_taxo <- read.delim("data/microbiome_dataSet1_taxonomy.tsv",row.names=1)
```

### Initial Filtering

Make sure all null (NA) elements are set to zero. Any all-zero rows and columns are removed.

```{r data_trns, include=FALSE}
#| echo: false
#| message: false
#| warning: false

#NAs to zeros
X[is.na(X)] <- 0

#remove all-zero samples
aux_allzero <- which(rowSums(X)==0)
if(length(aux_allzero)>0){
  X <- X[-aux_allzero,]
}

#remove all-zero taxa
aux_allzero <- which(colSums(X)==0)
if(length(aux_allzero)>0){
  X <- X[,-aux_allzero]
}

rm(aux_allzero)

if(any(X<0)){
  print("X has negative values. DO NOT CONTINUE.")
}

I <- dim(X)[1]; J <- dim(X)[2]
```

### Breakdown by study group and dysbiotic status

```{r EDA_table}
#| echo: false
#| message: false
#| warning: false

aux_table <- X_meta %>%
  count(diagnosis, dysbiotic) %>%
  # Reshape the data to have dysbiotic as columns (TRUE/FALSE)
  pivot_wider(names_from = dysbiotic, values_from = n, values_fill = 0) %>%
  # Rename columns to indicate TRUE/FALSE
  rename(Dysbiotic_True = `TRUE`, Dysbiotic_False = `FALSE`)

# Print the table
print(aux_table)
rm(aux_table)
```

### Sparsity and nonzero elements

Proportion of zeros in the matrix:

```{r EDA_sparsity_pre}
#| echo: false
#| message: false
#| warning: false

aux_sparsity <- round(length(which(X==0))/(I*J)*100,2)
print(paste0('the matrix is ',aux_sparsity,'% sparse before filtering rare taxa'))

rm(aux_sparsity)
```

Distribution/skew of nonzero elements:

```{r EDA_nonzero_pre}
#| echo: false
#| message: false
#| warning: false

hist(X[which(X>0)], breaks=50, main='Distribution of nonzero elements',xlab="reads per sample-OTU pair")
summary(X[which(X>0)])
```

### OTU analysis: rarity and number of species

Verify sequencing depth is the same across all samples:

```{r seqdepth_pre}
#| echo: false
#| message: false
#| warning: false

summary(rowSums(X))
```

OTU rarity: proportion of samples containing each of the species:

```{r OTU_rarity_pre}
#| echo: false
#| message: false
#| warning: false
aux_hist <- apply(X, 2, function(x) 100*length(which(x>0))/length(x))
hist(aux_hist, xlab='% samples with species',ylab='OTUs',main='OTU rarity',las=1,breaks=50)
summary(aux_hist)
rm(aux_hist)
```

Number of species per sample:

```{r species_pre}
#| echo: false
#| message: false
#| warning: false

hist(specnumber(X),breaks=50, xlab='OTUs',ylab='samples', main='number of species per sample')
summary(specnumber(X))
```

### Preprocessing

#### Binarization

Presence/absence matrix, basis for Bibit

$$
f(x) =
\begin{cases} 
1 & \text{if } x > 0, \\
0 & \text{otherwise.}
\end{cases}
$$

```{r trns_bin_pre}
#| echo: false
#| message: false
#| warning: false

Xbin <- X
Xbin[Xbin>0] <- 1
```

#### Square root transform

Variance stabilization, useful for factorization with NMF-NNSMR. **X** is backed up to **X**~orig~

```{r trns_sqrt_pre}
#| echo: false
#| message: false
#| warning: false
Xorig <- X
X <- sqrt(X)
Xsqrt <- X
```

### Heatmaps

Visualize the **X**, **X~orig~** and **X~sqrt~**

```{r plot_heatmaps_pre}
#| echo: false
#| message: false
#| warning: false
heatmap(Xorig, Rowv=NA, Colv=NA, scale="none", main="X", cexRow=0.5, cexCol=0.1)
heatmap(Xsqrt, Rowv=NA, Colv=NA, scale="none", main="X (sqrt transform)", cexRow=0.5, cexCol=0.1)
heatmap(Xbin, Rowv=NA, Colv=NA, scale="none", main="X (binarized)", cexRow=0.5, cexCol=0.1)
```

## 16S IBD - Filtered

If filtering is not applied, proceed directly to the [Hierarchical Clustering] section.

### Filtering rare taxa

Find and remove rare taxa: OTUs with only one count across all samples

```{r EDA_filter}
#| echo: false
#| message: false
#| warning: false

#restore raw data
X <- Xorig

aux_rare <- which(colSums(X)==1)
X <- X[,-aux_rare]

#update dimension variables
I <- dim(X)[1]; J <- dim(X)[2]

rm(aux_rare)
```

### Sparsity and nonzero elements

Proportion of zeros in the matrix:

```{r EDA_sparsity_post}
#| echo: false
#| message: false
#| warning: false

aux_sparsity <- round(length(which(X==0))/(I*J)*100,2)
print(paste0('the matrix is ',aux_sparsity,'% sparse after filtering rare taxa'))

rm(aux_sparsity)
```

Distribution/skew of nonzero elements:

```{r EDA_nonzero_post}
#| echo: false
#| message: false
#| warning: false

hist(X[which(X>0)], breaks=50, main='Distribution of nonzero elements',xlab="reads per sample-OTU pair")
summary(X[which(X>0)])
```

### OTU analysis: rarity and number of species

Verify changes in sequencing depth across all samples after filtering:

```{r seqdepth_post}
#| echo: false
#| message: false
#| warning: false

hist(rowSums(X),main='Sequencing depth after filtering',xlab='Reads per sample',ylab='samples',breaks=10)
summary(rowSums(X))
```

OTU rarity: proportion of samples containing each of the species:

```{r OTU_rarity_post}
#| echo: false
#| message: false
#| warning: false
aux_hist <- apply(X, 2, function(x) 100*length(which(x>0))/length(x))
hist(aux_hist, xlab='% samples with species',ylab='OTUs',main='OTU rarity after filtering',las=1,breaks=50)
summary(aux_hist)
rm(aux_hist)
```

Number of species per sample:

```{r species_post}
#| echo: false
#| message: false
#| warning: false

hist(specnumber(X),breaks=50, xlab='OTUs',ylab='samples', main='number of species per sample after filtering')
summary(specnumber(X))
```

### Preprocessing

#### Binarization

```{r trns_bin}
#| echo: false
#| message: false
#| warning: false

Xbin <- X
Xbin[Xbin>0] <- 1
```

#### Square root transform

```{r trns_sqrt}
#| echo: false
#| message: false
#| warning: false

Xorig <- X
X <- sqrt(X)
Xsqrt <- X
```

### Heatmaps

```{r plot_heatmaps}
#| echo: false
#| message: false
#| warning: false

heatmap(Xorig, Rowv=NA, Colv=NA, scale="none", main="X (filtered)", cexRow=0.5, cexCol=0.1)
heatmap(Xsqrt, Rowv=NA, Colv=NA, scale="none", main="X (filtered + sqrt transform)", cexRow=0.5, cexCol=0.1)
heatmap(Xbin, Rowv=NA, Colv=NA, scale="none", main="X (filtered + binarized)", cexRow=0.5, cexCol=0.1)
```

# Hierarchical Clustering

Perform hierarchical clustering on the samples using the Euclidean distance and average linkage across all features.

```{r hc}
#| echo: false
#| message: false
#| warning: false
d_IBD <- dist(Xorig,method = "euclidean")
hc_IBD <- hclust(d_IBD, method = "average")
plot(hc_IBD,cex=0.5)
clusterCut <- cutree(hc_IBD, 9)
print('study groups by cluster')
table(clusterCut, X_meta$diagnosis)
print('----------------------------------------')
print('dysbiotic status by cluster')
table(clusterCut, X_meta$dysbiotic)

```

### Check inter-sample distance and similarity

Manually inspect the computed distances and Jaccard Index (IoU) for any pair of samples by sample name Using the dendrogram above, compare the distance and similarity between any two samples.

```{r hc_inspect_samples}
#| echo: false
#| message: false
#| warning: false

s1 <- X['M2042C18',] #sample 1
s2 <- X['H4023C13',] #sample 2
as.matrix(dist(rbind(s1,s2))) #display computed distance

#get nonzero features in each sample
s1_t <- which(s1!=0)
s2_t <- which(s2!=0)
#calculate the intersection over union
aux_iou <- length(intersect(s1_t,s2_t))/length(union(s1_t,s2_t))
print(paste0('IoU_cols =',round(aux_iou,2)))

```

### Inspect and plot clusters

Presence/absence heatmaps to illustrate the sparsity of the resulting clusters. Text output indicates:

-   Number of features present in all samples in the cluster

-   Number of features containing no counts in any of the samples in the cluster.

```{r hc_inspect_clusters}
#| echo: false
#| message: false
#| warning: false

#empty object to store 'orphan' samples (not clustered)
aux_1sample <- c()

for(bcc in unique(clusterCut)){
  aux_samples <- clusterCut[clusterCut==bcc]
  if(length(aux_samples)>1){
    aux_C <- Xbin[names(aux_samples),]
    heatmap(aux_C, Rowv=NA, Colv=NA, scale="none", main=paste0("HC",bcc), cexRow=0.5, cexCol=0.1)
    
    print(paste0('Cluster ',bcc,':'))
    print(paste0(length(which(colSums(aux_C)==nrow(aux_C))),' features have counts on all ',nrow(aux_C),' samples'))
    print(paste0(length(which(colSums(aux_C)==0)),' features have no counts in any of the ',nrow(aux_C),' samples'))
    print('')
    print('features present in ___ samples')
    print(table(colSums(aux_C)))
    
  }
  else{
    print(paste('Cluster',bcc,'has only one sample'))
    aux_1sample <- c(aux_1sample, names(clusterCut[clusterCut==bcc]))
  }
  print('----------------------------------------')
}





print('')
print('samples in single-sample clusters')
aux_1sample
```

Visualize all single-sample clusters in a heatmap to illustrate their individual sparsity.

```{r hc_1sample}
#| echo: false
#| message: false
#| warning: false

aux_1sample_m <- Xbin[aux_1sample,]
heatmap(aux_1sample_m, Rowv=NA, Colv=NA, scale="none", main='single-sample clusters', cexRow=0.5, cexCol=0.1)

as.matrix(dist(aux_1sample_m))

for(cc in aux_1sample){
  print(paste('sample',cc,'has',length(which(Xbin[cc,]==0)),'taxa without reads'))
}
```

### 

### Inspect a single cluster

Create a temporary object with the bicluster, selected by number. Verify distances between its samples.

```{r hc_inspect_single}
#| echo: false
#| message: false
#| warning: false

cc <- 3
aux_c <- Xorig[names(clusterCut[clusterCut==cc]),]
as.matrix(dist(aux_c))
```

```{r hc_inspect_meta}
#| echo: false
#| message: false
#| warning: false

#input a study group and a dysbiotic status of interest
aux_diag <- 'IBD_CD'
aux_dysb <- 'T'

#number of results for median taxonomic abundance
aux_n <- 10

#convert TRUE and FALSE to T and F
X_meta$dysbiotic <- ifelse(X_meta$dysbiotic,'T','F')

#get row names of all samples and of nonzero taxa in the cluster
aux_samples <- names(clusterCut[clusterCut==cc])
aux_taxa <- which(colSums(aux_c)!=0)

#get the metadata and taxonomy of the clustered rows and the nonzero taxa
X_meta_c <- X_meta[aux_samples,]
X_taxo_c <- X_taxo[aux_taxa,]

#visualize the group distribution of the selection of samples
as.data.frame(table(X_meta_c$diagnosis)) #samples by diagnosis
as.data.frame(table(X_meta_c$dysbiotic)) #samples by dysbiotic status
as.data.frame(table(X_taxo_c$class)) #taxa by class
as.data.frame(table(X_taxo_c$family)) #taxa by family

#select a diagnosis and get a submatrix from the bicluster
X_meta_c_diag <- X_meta_c[X_meta_c$diagnosis==aux_diag,]
aux_c_diag <- aux_c[rownames(X_meta_c_diag),]

#select by dysbiotic status and get a submatrix from the bicluster
X_meta_c_dysb <- X_meta_c[X_meta_c$dysbiotic==aux_dysb,]
aux_c_dysb <- aux_c[rownames(X_meta_c_dysb),]

#visualize the selections to find patterns
heatmap(aux_c_diag, Rowv=NA, Colv=NA, scale="none", main=paste0('Cluste r',cc,' (',aux_diag,')'),cexRow=1,cexCol=0.25)

heatmap(aux_c_dysb, Rowv=NA, Colv=NA, scale="none", main=paste0('Cluster ',cc,' (',
                                                                ifelse(aux_dysb=='T','','non-'),
                                                                'dysbiotic)'),cexRow=1,cexCol=0.25)

#keep only nonzero taxa
aux_c_diag <- aux_c_diag[,aux_taxa]
aux_c_dysb <- aux_c_dysb[,aux_taxa]

#get the median abundance of taxa by diagnosis and by dysbiotic status
#by diagnostic
aux_c_diag_colsum_median <- apply(aux_c_diag,2,median)
print(paste('Top',aux_n,'median abundance taxa in',paste('Cluster',cc),aux_diag,':'))
head(sort(aux_c_diag_colsum_median[aux_c_diag_colsum_median>0], decreasing = TRUE),n=aux_n)
print('----------------------------------------')
print(paste('Bottom',aux_n,'median abundance taxa in',paste('Cluster',cc),aux_diag,':'))
head(sort(aux_c_diag_colsum_median[aux_c_diag_colsum_median>0], decreasing = FALSE),n=aux_n)

print('========================================')
#by dysbiosis
aux_c_dysb_colsum_median <- apply(aux_c_dysb,2,median)
print(paste0('Top',aux_n,'median abundance taxa in ',paste('Cluster',cc),ifelse(aux_dysb=='T','','non-'),'dysbiotic:'))
head(sort(aux_c_dysb_colsum_median[aux_c_dysb_colsum_median>0], decreasing = TRUE),n=aux_n)
print('----------------------------------------')
print(paste0('Bottom',aux_n,'median abundance taxa in ',paste('Cluster',cc),ifelse(aux_dysb=='T','','non-'),'dysbiotic:'))
head(sort(aux_c_dysb_colsum_median[aux_c_dysb_colsum_median>0], decreasing = FALSE),n=aux_n)

```

```         
```
